Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeast Taiwan

### 2019-10-08 analysis/paper/paper.Rmd
added bibtex, and revised introduction and method section

Culture contact between indigenous people and colonizer with imperial power usually leads to profound changes in local indigenous societies. We explore the culture changes in indigenous societies in northeast Taiwan that experienced foreign contacts with the Europeans and Chinese since 17th century by examining the local pottery production. Pottery production can reflect prehistoric socioeconomic patterns since it relates to not only economic base for households but also societies as whole when specialized production was organized that hints changes in production systems. Based on the assumption that the specialized mass production will lead to uniformity of the product due to increased skills, routinization, and lower diversity of producers, we may be able to determine the type of production, and the presence of elite control, from artefact analysis (Arnold 2000: 334; Blackman, et al. 1993; Costin 1991; Hirshman, et al. 2010).

We observed pottery production in northeast Taiwan in the late Iron Age period and historical period (1400-1900 AD) presenting a high consistency in form and shape compared to other pottery found from other regions on this island that hints the emergence of pottery specialization. We asked the questions: Did foreign contact have an impact on indigenous pottery production in Northeast Taiwan that can be detected in the shape of the vessels? Did the pottery production become more standardized and homogeneous in shape after foreign contact? Did the contact with European cause more homogeneous shape than Chinese contact period because of large scale exchange network?

Pottery specialization can be identified by examining the standardization of pottery measurements (citation). However, traditional typological and linear measurements are limited because they can be insensitive to subtle variations resulting from changes in craft specialization. Taking a new approach to the measurement of craft specialization, we studied shape and standardization of locally made pottery to identify changes in pottery production at Kiwulan (1400-1900 AD), a large multi-component archaeological site in northeast Taiwan. We apply geometric morphometric methods to study artifact shapes to investigate if there are any changes resulting from foreign contact of European and Chinese that might indicate social changes in the indigenous society.

@@

# Methods

@@

To prepare the data for shape analysis, we first traced the outlines of pottery from previously outline drawings done by excavation team using Inkscape software. The digitised outlines were converted into a list of x-y pixel coordinates for elliptic Fourier analysis (EFA), which can assess pottery morphological difference among pottery outlines from three occupation contexts. The EFA generates harmonic coefficients were analysed by principal component analysis (PCA) to identity major patterns in data through dimensionality reduction. The principal components (PCs) were analysed with a multivariate analysis of variance (MANOVA) to test for significant effects of the occupation context on shape variances. Finally, we computed a significance test for the equality of coefficients of variation (CVs) among multiple groups of PCs to understand the differences in shape standardisation. Outline analyses were carried out using the package Momocs in the software R.

>>>>

Ceramic production can reflect prehistoric socioeconomic patterns since it relates to not only economic base for households but also societies as a whole when specialized production was organized. Identifying the specialization in ceramic production will help to understand how complex societies form and further explore the underlying mechanism. One way to examine ceramic specialization is to identify whether ceramics present standardization in production, which based on the assumption that the specialized mass production will lead to uniformity of the product due to increased skills, routinization, and lower diversity of producers that enable us to interpret social organization or social relations such as type of production, and the presence of elite control [@Arnold2000, pp. 334; @Stark1995; @Costin1991].

Several variables have been proposed to measure ceramic standardization, including metric, compositional, and technological attributes [@Rice1991; @Blackman1993; @Costin1991; @Hirshman2010; @Tite1999]. However, traditional typological and linear measurements are limited because they can be insensitive to subtle variations resulting from changes in craft specialization.  Taking a new approach to the measurement of craft specialization, we studied shape and standardization of locally made pottery to identify changes in pottery production at Kiwulan (1400-1900 AD), a large multi-component archaeological site in northeast Taiwan. We apply geometric morphometric methods to study artifact shapes to investigate if there are any changes resulting from foreign contact of European and Chinese that might indicate social changes in the indigenous society.

Culture contact between indigenous people and colonizer with imperial power usually leads to profound changes in local indigenous societies. We explore the culture changes in indigenous societies in northeast Taiwan that experienced foreign contacts with the Europeans and Chinese since 17th century by examining the local pottery production. We observed pottery production in northeast Taiwan in the late Iron Age period and historical period (1400-1900 AD) presenting a high consistency in form and shape compared to other pottery found from other regions on this island that hints the emergence of pottery specialization. We asked the questions: Did foreign contact have an impact on indigenous pottery production in Northeast Taiwan that can be detected in the shape of the vessels? Did the pottery production become more standardized and homogeneous in shape after foreign contact? Did the contact with European cause more homogeneous shape than Chinese contact period because of large scale exchange network?

@@

# Materials and methods

## Archaeological pottery from Kiwulan

Ceramics data analysed in this paper come from the Upper Culture Layer of Kiwulan, northeast Taiwan. Those ceramics dated from AD 1400 to AD 1900, which covers 500 years from the late Iron Age to the historical period with the onset of the European presence in the 17th century. Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides. The Upper Culture Layer experienced frequent foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since 19th century. The excavation revealed abundant artifacts in which potsherds are the dominate materials throughout the site. Imported ceramics from mainland China, stonewares, and ornament elements such as beads were usually found in the Upper Culture Layer that indicates the prosperous international trading activities in the 17th century during the European presence in Taiwan. In addition to artifacts, features were also found, such as burials, middens, and postholes with in-situ posts that explains Kiwulan was a large settlement site.

@@

The drawings of pottery with half cross-section from rim to body were imported into Inkscape software for outline digitalization. All half cross-section images were copied and flipped to create enclosed images in order to be ran in the R environment. The digitised outlines were converted into a list of x-y pixel coordinates for elliptic Fourier analysis (EFA), which can assess pottery morphological difference among pottery outlines from three occupation contexts. The EFA generates harmonic coefficients were analysed by principal component analysis (PCA) to identity major patterns in data through dimensionality reduction. The principal components (PCs) were analysed with a multivariate analysis of variance (MANOVA) to test for significant effects of the occupation context on shape variances. Finally, we computed a significance test for the equality of coefficients of variation (CVs) among multiple groups of PCs to understand the differences in shape standardisation. Outline analyses were carried out using the package Momocs in the software R.
